---
title: "2024 MD eviction analysis"
author: "Sandhya Kajeepeta"
date: "2024-12-09"
output: html_document
---

```{r load packages,echo=FALSE, results=FALSE, message=FALSE, warning=FALSE}
library(tidyverse)
library(knitr)
library(kableExtra)
library(ggplot2)
library(sf)
library(readxl)
library(MASS)
library(pscl)
```

# Executed evictions
## Total executed evictions
```{r load eviction data, echo=FALSE, warning=FALSE, message=FALSE}
evic <- read_csv("./2024 data/District_Court_of_Maryland_Eviction_Case_Data.csv")
#Source: https://app.powerbigov.us/view?r=eyJrIjoiYWI1Yzg0YjYtNDFkZS00MDUyLThlMDctYmE1ZjY5MGI0MWJhIiwidCI6IjdkM2I4ZDAwLWY5YmUtNDZlNy05NDYwLTRlZjJkOGY3MzE0OSJ9&pageName=ReportSection 

#convert dates to datetime vars
evic$`Event Date` <- as.POSIXct(evic$`Event Date`, format="%m/%d/%Y")
evic$`Evicted Date` <- as.POSIXct(evic$`Evicted Date`, format="%m/%d/%Y")

#restrict to evictions in Oct 2023-Sep 2024 and group by zip code
#Time period = Oct 2023 - Sep 2024
df1 <- evic %>% filter(`Event Type`=="Warrant of Restitution - Return of Service - Evicted" & `Event Date`>"2023-9-30" & `Event Date`<"2024-10-01") %>% group_by(`Tenant ZIP Code`) %>% summarize(count=n())
colnames(df1) <- c("ZIP", "evictions")
```

```{r load housing data by zip, echo=FALSE, warning=FALSE, message=FALSE}
acs <- read_csv("./Data/ACS_5Y_2022.csv")
#Source: ACS 5-year estimates 2022: https://data.census.gov/table/ACSST5Y2022.S2502?q=S2502&g=040XX00US24$8600000&tp=true 

#select number of renter-occupied housing units and number of Black renter-occupied housing units
acs <- acs %>% dplyr::select(NAME, S2502_C05_001E, S2502_C05_003E)

#clean up vars
colnames(acs) <- c("ZIP","renter_units","black_renter_units")
acs$ZIP <- as.numeric(substring(acs$ZIP, 7, 11))
```

```{r join data, echo=FALSE, warning=FALSE, message=FALSE}
df1 <- acs %>% left_join(df1, by="ZIP")

#drop zip codes with 0 rental units or missing ACS data
df1 <- df1 %>% filter(renter_units > 0)

#if no eviction data, assume that there were 0 evictions
df1$evictions <- ifelse(is.na(df1$evictions), 0, df1$evictions)

#create vars of interest
df1$perc_black_rent <- df1$black_renter_units/df1$renter_units*10
df1$evic_rate <- df1$evictions/df1$renter_units*100
```

```{r plot vars zip, echo=FALSE, results=FALSE, warning=FALSE, fig.show='hide'}
plot <-ggplot(df1, mapping=aes(x=perc_black_rent, y=evic_rate)) + geom_point()
plot

hist(df1$evic_rate)
hist(df1$perc_black_rent)

#drop zip codes with fewer than 50 rental units
df1 <- df1 %>% filter(renter_units >= 50)
```

```{r poisson or neg binomial zip, echo=FALSE, results=FALSE, warning=FALSE, fig.show='hide'}
p_model <- glm(formula=evictions~perc_black_rent + offset(log(renter_units)), data=df1, family="poisson")
nb_model <- glm.nb(formula=evictions~perc_black_rent + offset(log(renter_units)), data=df1)

#Residual plot for Poisson regression
p_res <- resid(p_model)
plot(fitted(p_model), p_res, col='steelblue', pch=16,
     xlab='Predicted Offers', ylab='Standardized Residuals', main='Poisson')
abline(0,0)

#Residual plot for negative binomial regression 
nb_res <- resid(nb_model)
plot(fitted(nb_model), nb_res, col='steelblue', pch=16,
     xlab='Predicted Offers', ylab='Standardized Residuals', main='Negative Binomial')
abline(0,0)

#Residuals for negative binomial plot are much smaller, Poisson residuals extend all the way to 25 - suggests that negative binomial is a better fit than Poisson

pchisq(2 * (logLik(nb_model) - logLik(p_model)), df = 1, lower.tail = FALSE)

#Negative binomial offers significantly better fit
```
<iframe title="Rate of executed evictions by ZIP code (Oct 2023 - Sep 2024)" aria-label="Map" id="datawrapper-chart-jdVQK" src="https://datawrapper.dwcdn.net/jdVQK/1/" scrolling="no" frameborder="0" style="width: 0; min-width: 100% !important; border: none;" height="485" data-external="1"></iframe><script type="text/javascript">!function(){"use strict";window.addEventListener("message",(function(a){if(void 0!==a.data["datawrapper-height"]){var e=document.querySelectorAll("iframe");for(var t in a.data["datawrapper-height"])for(var r=0;r<e.length;r++)if(e[r].contentWindow===a.source){var i=a.data["datawrapper-height"][t]+"px";e[r].style.height=i}}}))}();
</script>

<iframe title="Black renters as percentage of renter population by ZIP code (2022)" aria-label="Map" id="datawrapper-chart-wE0kS" src="https://datawrapper.dwcdn.net/wE0kS/1/" scrolling="no" frameborder="0" style="width: 0; min-width: 100% !important; border: none;" height="499" data-external="1"></iframe><script type="text/javascript">!function(){"use strict";window.addEventListener("message",(function(a){if(void 0!==a.data["datawrapper-height"]){var e=document.querySelectorAll("iframe");for(var t in a.data["datawrapper-height"])for(var r=0;r<e.length;r++)if(e[r].contentWindow===a.source){var i=a.data["datawrapper-height"][t]+"px";e[r].style.height=i}}}))}();
</script>

```{r analysis zip, echo=FALSE, warning=FALSE, message=FALSE}
m <- glm.nb(formula=evictions~perc_black_rent + offset(log(renter_units)), data=df1)
print("Eviction rate")
est <- exp(cbind(Estimate = coef(m), confint(m)))
kable(est, "pipe")

```
For ZIP codes in Maryland with 50 or more rental units, ZIP codes with a higher percentage of Black renters face a higher rate of total evictions.

A 10-percentage-point increase in the percent of Black renters in a ZIP code is associated with a 14% increase in the rate of evictions.

## Executed holdover evictions
```{r holdover eviction data, echo=FALSE, warning=FALSE, message=FALSE}
#restrict to evictions in Oct 2023 to Sep 2024, restrict to holdover evics, and group by zip code
#Time period = Oct 2023 - Sep 2024
df2 <- evic %>% filter(`Event Type`=="Warrant of Restitution - Return of Service - Evicted" & `Event Date`>"2023-9-30" & `Event Date`<"2024-10-01"& `Case Type`=="Tenant Holding Over") %>% group_by(`Tenant ZIP Code`) %>% summarize(count=n())
colnames(df2) <- c("ZIP", "holdover_evictions")
```

```{r join holdover data, echo=FALSE, warning=FALSE, message=FALSE}
df2 <- acs %>% left_join(df2, by="ZIP")

#drop zip codes with 0 rental units or missing ACS data
df2 <- df2 %>% filter(renter_units > 0)

#if no eviction data, assume that there were 0 evictions
df2$holdover_evictions <- ifelse(is.na(df2$holdover_evictions), 0, df2$holdover_evictions)

#create vars of interest
df2$perc_black_rent <- df2$black_renter_units/df2$renter_units*10
df2$evic_rate <- df2$holdover_evictions/df2$renter_units*100
```

```{r plot vars zip holdover, echo=FALSE, results=FALSE, warning=FALSE, fig.show='hide'}
plot <-ggplot(df2, mapping=aes(x=perc_black_rent, y=evic_rate)) + geom_point()
plot

hist(df2$evic_rate)
hist(df2$perc_black_rent)

#drop zip codes with fewer than 50 rental units
df2 <- df2 %>% filter(renter_units >= 50)
```

```{r poisson or neg binomial zip holdover, echo=FALSE, results=FALSE, warning=FALSE, fig.show='hide'}
p_model <- glm(formula=holdover_evictions~perc_black_rent + offset(log(renter_units)), data=df2, family="poisson")
nb_model <- glm.nb(formula=holdover_evictions~perc_black_rent + offset(log(renter_units)), data=df2)

#Residual plot for Poisson regression
p_res <- resid(p_model)
plot(fitted(p_model), p_res, col='steelblue', pch=16,
     xlab='Predicted Offers', ylab='Standardized Residuals', main='Poisson')
abline(0,0)

#Residual plot for negative binomial regression 
nb_res <- resid(nb_model)
plot(fitted(nb_model), nb_res, col='steelblue', pch=16,
     xlab='Predicted Offers', ylab='Standardized Residuals', main='Negative Binomial')
abline(0,0)

#Residuals for negative binomial plot are smaller - suggests that negative binomial is a better fit than Poisson

pchisq(2 * (logLik(nb_model) - logLik(p_model)), df = 1, lower.tail = FALSE)

#Negative binomial offers significantly better fit
```

```{r analysis zip holdover, echo=FALSE, warning=FALSE, message=FALSE}
m <- glm.nb(formula=holdover_evictions~perc_black_rent + offset(log(renter_units)), data=df2)
print("Holdover eviction rate")
est <- exp(cbind(Estimate = coef(m), confint(m)))
kable(est, "pipe")

```
For ZIP codes in Maryland with 50 or more rental units, ZIP codes with a higher percentage of Black renters face a higher rate of holdover evictions.

A 10-percentage-point increase in the percent of Black renters in a ZIP code is associated with a 10% increase in the rate of holdover evictions, but this effect is not statistically significant.

# Warrants of restitution
## Total warrants of restitution

```{r warrant data, echo=FALSE, warning=FALSE, message=FALSE}
#restrict to warrants in Oct 2023 to Sep 2024 and group by zip code
#Time period = Oct 2023 - Sep 2024
df3 <- evic %>% filter(`Event Date`>"2023-9-30" & `Event Date`<"2024-10-01") %>% group_by(`Tenant ZIP Code`) %>% summarize(count=n())
colnames(df3) <- c("ZIP", "warrants")
```

```{r join warrant data, echo=FALSE, warning=FALSE, message=FALSE}
df3 <- acs %>% left_join(df3, by="ZIP")

#drop zip codes with 0 rental units or missing ACS data
df3 <- df3 %>% filter(renter_units > 0)

#if no eviction data, assume that there were 0 warrants
df3$warrants <- ifelse(is.na(df3$warrants), 0, df3$warrants)

#create vars of interest
df3$perc_black_rent <- df3$black_renter_units/df3$renter_units*10
df3$evic_rate <- df3$warrants/df3$renter_units*100
```

```{r plot vars zip warrant, echo=FALSE, results=FALSE, warning=FALSE, fig.show='hide'}
plot <-ggplot(df3, mapping=aes(x=perc_black_rent, y=evic_rate)) + geom_point()
plot

hist(df3$evic_rate)
hist(df3$perc_black_rent)

#drop zip codes with fewer than 50 rental units
df3 <- df3 %>% filter(renter_units >= 50)
```

```{r poisson or neg binomial zip warrant, echo=FALSE, results=FALSE, warning=FALSE, fig.show='hide'}
p_model <- glm(formula=warrants~perc_black_rent + offset(log(renter_units)), data=df3, family="poisson")
nb_model <- glm.nb(formula=warrants~perc_black_rent + offset(log(renter_units)), data=df3)

#Residual plot for Poisson regression
p_res <- resid(p_model)
plot(fitted(p_model), p_res, col='steelblue', pch=16,
     xlab='Predicted Offers', ylab='Standardized Residuals', main='Poisson')
abline(0,0)

#Residual plot for negative binomial regression 
nb_res <- resid(nb_model)
plot(fitted(nb_model), nb_res, col='steelblue', pch=16,
     xlab='Predicted Offers', ylab='Standardized Residuals', main='Negative Binomial')
abline(0,0)

#Residuals for negative binomial plot are smaller, Poisson residuals extend all the way to 60 - suggests that negative binomial is a better fit than Poisson

pchisq(2 * (logLik(nb_model) - logLik(p_model)), df = 1, lower.tail = FALSE)

#Negative binomial offers significantly better fit
```

```{r analysis zip warrant, echo=FALSE, warning=FALSE, message=FALSE}
m <- glm.nb(formula=warrants~perc_black_rent + offset(log(renter_units)), data=df3)
print("Warrants of restitution filing rate")
est <- exp(cbind(Estimate = coef(m), confint(m)))
kable(est, "pipe")

```
For ZIP codes in Maryland with 50 or more rental units, ZIP codes with a higher percentage of Black renters face a higher rate of filings of warrants of restitution.

A 10-percentage-point increase in the percent of Black renters in a ZIP code is associated with a 32% increase in the rate of filings of warrants of restitution.

## Warrants of restitution for holdover evictions
```{r holdover warrant data, echo=FALSE, warning=FALSE, message=FALSE}
#restrict to warrants in Oct 2023 to Sep 2024, restrict to holdover, and group by zip code
#Time period = Oct 2023 - Sep 2024
df4 <- evic %>% filter(`Event Date`>"2023-9-30" & `Event Date`<"2024-10-01" & `Case Type`=="Tenant Holding Over") %>% group_by(`Tenant ZIP Code`) %>% summarize(count=n())
colnames(df4) <- c("ZIP", "holdover_warrants")
```

```{r join holdover warrant data, echo=FALSE, warning=FALSE, message=FALSE}
df4 <- acs %>% left_join(df4, by="ZIP")

#drop zip codes with 0 rental units or missing ACS data
df4 <- df4 %>% filter(renter_units > 0)

#if no eviction data, assume that there were 0 holdover warrants
df4$holdover_warrants <- ifelse(is.na(df4$holdover_warrants), 0, df4$holdover_warrants)

#create vars of interest
df4$perc_black_rent <- df4$black_renter_units/df4$renter_units*10
df4$evic_rate <- df4$holdover_warrants/df4$renter_units*100
```

```{r plot vars zip holdover warrant, echo=FALSE, results=FALSE, warning=FALSE, fig.show='hide'}
plot <-ggplot(df4, mapping=aes(x=perc_black_rent, y=evic_rate)) + geom_point()
plot

hist(df4$evic_rate)
hist(df4$perc_black_rent)

#drop zip codes with fewer than 50 rental units
df4 <- df4 %>% filter(renter_units >= 50)
```

```{r poisson or neg binomial zip holdover warrant, echo=FALSE, results=FALSE, warning=FALSE, fig.show='hide'}
p_model <- glm(formula=holdover_warrants~perc_black_rent + offset(log(renter_units)), data=df4, family="poisson")
nb_model <- glm.nb(formula=holdover_warrants~perc_black_rent + offset(log(renter_units)), data=df4)

#Residual plot for Poisson regression
p_res <- resid(p_model)
plot(fitted(p_model), p_res, col='steelblue', pch=16,
     xlab='Predicted Offers', ylab='Standardized Residuals', main='Poisson')
abline(0,0)

#Residual plot for negative binomial regression 
nb_res <- resid(nb_model)
plot(fitted(nb_model), nb_res, col='steelblue', pch=16,
     xlab='Predicted Offers', ylab='Standardized Residuals', main='Negative Binomial')
abline(0,0)

#Residuals for negative binomial plot are smaller, Poisson residuals extend all the way to 10 - suggests that negative binomial is a better fit than Poisson

pchisq(2 * (logLik(nb_model) - logLik(p_model)), df = 1, lower.tail = FALSE)

#Negative binomial offers significantly better fit
```

```{r analysis zip warrant holdover, echo=FALSE, warning=FALSE, message=FALSE}
m <- glm.nb(formula=holdover_warrants~perc_black_rent + offset(log(renter_units)), data=df4)
print("Warrants of restitution filing rate for holdover evictions")
est <- exp(cbind(Estimate = coef(m), confint(m)))
kable(est, "pipe")

```
For ZIP codes in Maryland with 50 or more rental units, ZIP codes with a higher percentage of Black renters face a higher rate of filings of warrants of restitution for holdover evictions.

A 10-percentage-point increase in the percent of Black renters in a ZIP code is associated with a 13% increase in the rate of filings of warrants of restitution for holdover evictions.

```{r exports datasets for maps, echo=FALSE, warning=FALSE, message=FALSE}
write.csv(df1, "./Map Data/2024 map data/total executed evictions.csv")
write.csv(df2, "./Map Data/2024 map data/executed holdover evictions.csv")
write.csv(df3, "./Map Data/2024 map data/warrants of restitution.csv")
write.csv(df4, "./Map Data/2024 map data/holdover warrants.csv")

```
